Computer-Assisted Analysis of the \\ Anderson-Hájek Ontological Controversy

### 2014-10-24 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
Add ToDo
Add missing bibliography items

\noindent The axioms in G\"odel's ontological proof
\citep{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail
what is called \emph{modal collapse}
\citep{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which
has recently been confirmed with higher-order automated theorem
provers \citep{C40,J30}, has led to strong criticism of the argument
and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} suggested that
greater care is necessary in the semantics of higher-order quantifiers
in the presence of modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}, \citet{Hajek2002}
and Bjordal \citet{bjordal99}, proposed slight emendations of G\"odel's
axioms and definitions. They require neither comprehension
restrictions nor more complex semantics. Therefore, they are
technically simpler to analyze with computer support. We have
formalized them using the proof assistant Isabelle/HOL \citep{Isabelle}
together with the automated higher-order reasoners Leo-II \citep{C26},
Satallax \citep{brown2012satallax}, Metis 
\citep{Hurd03first-orderproof}, and Nitpick \citep{Nitpick}.   Our
formalizations\footnote{The   formalizations are available in the
subdirectories \url{Anderson}, \url{Hájek} and \url{Bjordal} at \url{gi
thub.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/
}.} employ the embedding of higher-order modal logic (HOML) in
classical higher-order logic (HOL) as introduced in previous work
\citep{C40,J30,J23}. We explored the effect of different domain
conditions on the provability of lemmas, theorems and even axioms.
This was motivated by a controversy between Hájek and Anderson
regarding the redundancy of some axioms in Anderson's emendation. In
\emph{constant domain semantics}, the individual domains are the same
in all possible worlds. In \emph{varying domain semantics}, the
domains may vary from world to world. This variation is technically
encoded with the help of an existence relation expressing which
individuals actually exist in each world. Quantifiers are then
uniformly formalized as \emph{actualistic quantifiers} (i.e. guarded by
the existence relation). Our main results are summarized here.

@@

@@

The redundancy of A4 and A5 is particularly controversial. Magari \cite{Magari1988} claimed that A4 and A5 are superfluous\footnote{
  An axiom $A$ is superfluous w.r.t. a set of axioms $S$ iff T3 is entailed by $S \setminus \{ A \}$.
}, arguing that T3 is true in all models of the other axioms and definitions by Gödel. \citet[p.~5-6]{Hajek_Magari_and_others_1996} investigated this further, refuted Magari's claim, but claimed that A4 and A5 are indeed superfluous under the assumption of an additional axiom (PEP, pointwise equality for positiveness), which Magari probably assumed silently. Moreover, Hájek \citet[p.~2]{Hajek_Magari_and_others_1996} cites his earlier work\footnote{Although \citep{Hajek_der_Mathematiker_2001} precedes \citep{Hajek_Magari_and_others_1996} in writing, it was published only 5 years later, in German.} \citep{Hajek_der_Mathematiker_2001}, where he claims (in Theorem 5.3) that for Anderson's emended theory \citep{anderson90:_some_emend_of_goedel_ontol_proof}, A4 and A5 are not only superfluous, but also redundant. \citet[footnote 1 in p.~1]{AG1996, footnote 1, page 1}, in a footnote, rebutted Hájek's claim, arguing that the redundancy of A4 and A5 holds only under constant domain semantics, while Anderson's emended theory ought to be taken under Cocchiarella's semantics \citep{Cocchiarella} (a varying domain semantics). Our results show that Hájek was originally right, under both constant and varying domain semantics.

@@

Additionally, Anderson's 14th footnote \citep[footnote 14]{anderson90:_some_emend_of_goedel_ontol_proof} remarks that only the quantifiers in T3' and in A:D2 need to be interpreted as actualistic quantifiers, while others may be taken as possibilistic quantifiers. Our computer-assisted study of this mixed variant shows that A4 is still redundant, but A5 becomes independent (hence not redundant). However, T3 becomes unprovable.

\section{Bjordal's Emendation (cf. Appendix \ref{apx:Bjordal})}

For both \textbf{constant domain semantics} and \textbf{varying domain semantics}, the following results hold:

    G\"odel's axiom
  A2, A3 can be quickly automatically derived in logic \K from
  Bjordal's definition B:D. A4 can be proved in logic \KT
  (reflexivity). Proving G\"odel's D1 from B:D is possible in logic
  \KFour. Conversely, the proof that B:D follows from D1, A2, A3 and
  A4 is possible already in logic \K. Hence, Bjordal's lemma
  B:L1 holds in logic \SFour. The provers also show that theorem T3
  follows from B:D, B:A1 and B:A2 already in logic \KB. Modal collapse
  does not follow in Bjordal's setting as Nitpick demonstrates with a
  countermodel (consisting of two worlds and one individual).

  \BWP{@Bruno: we could eventually also
  also say that for MC1 we need a bigger countermodel?}

\BWP{ToDo: discuss the fact that Bjordal's proof is more resistant to MC than Anderson's}

\section{Conclusions}

Anderson's emendation (cf. Appendix \ref{apx:Anderson}; ,
which we have analysed for different domain conditions. These
variations were motivated by various comments on Anderson's work in
the literature.

In this approach full comprehension is
naturally ``built-in'' since the underlying HOL supports
$\lambda$-abstraction.

Summary (what else can we say here, feel free to add): Using our approach, the formalization and (partly) automated
analysis of different variants of Anderson's and Bjordal's emendations
of G\"odel's ontological argument has been surprisingly
straightforward. The provers confirmed the claimed results and in a
few cases they have even contributed some novel insights. The
weakening of the comprehension principles would clearly constitute
another interesting parameter for further experiments. However, this
seems hard to achieve in our approach, since full comprehension is
naturally built-in.

%\bibliographystyle{plain}
%\bibliography{Bibliography}

@@

\BWP{I think we need more compact presentation of the proofs.}

@@

% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
In Bjordal's emendation $G$ (God-like) is taken as primitive and $P$ (Positive) is defined (cf. definition D).
\begin{itemize}
\item[B:D] A formulas $\phi$ is positive iff it is necessarily the case
  that anything which is God-like has the property $\phi$.
  $$\hol{P(\phi) \equiv \nec \allq x (G(x) \imp \phi(x))}$$ 
\item[B:L1] D is logically equivalent in S4 with the union of G\"odel's definition D1 and axioms A2, A3 and A4.
  $$\hol{D \biimp D1 \wedge A2 \wedge A3 \wedge A4}$$
  The proof splits into the two implication directions B:L1$^\rightarrow$ and B:L1$^\leftarrow$. B:L1$^\rightarrow$ can be further split into four single steps.
\item[B:D2] $\phi$ is a maximal composite of object $x$'s positive properties iff $x$ has $\phi$ and $\phi$ is positive and all positive properties $\psi$ which $x$ has are such that is necessarily the case that all objects which have $\phi$ also have $\psi$.
  $$\hol{MCP(\phi,x) \equiv (\phi(x) \wedge P(\phi)) \wedge \allq \psi ((\psi(x) \wedge P(\psi)) \imp \nec \allq y (\phi(y) \imp \psi(y)))}$$
\item[B:D3] $x$ has the $N$-property iff x is such that if $\phi$ is a maximal composite of $x$'s positive properties then it is necessary that some object $y$ has the property $\phi$.
  $$\hol{N(x) \equiv \allq \phi (MCP(\phi,x) \imp \nec \allq y \phi(y))}$$
\item[B:A1] If a property is positive, its negation is not positive:
  $$\hol{\allq \varphi [P(\varphi) \imp \neg P(\neg \varphi)]}$$ 
\item[B:A2] The $N$-property is positive.
 $$\hol{P(N)}$$
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Frode's Alternative \cite{ToDo:Frode}.\label{fig:frode}} 
% \end{figure}

>>>>

\noindent The axioms in G\"odel's ontological proof
\citep{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) entail
what is called \emph{modal collapse}
\citep{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended. This fact, which
has recently been confirmed with higher-order automated theorem
provers \citep{C40,J30}, has led to strong criticism of the argument
and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took
greater care of the semantics of higher-order quantifiers
in the presence of modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof}, \citet{Hajek2002}
and \citet{bjordal99}, proposed emendations of G\"odel's
axioms and definitions. They require neither comprehension
restrictions nor more complex semantics. Therefore, they are
technically simpler to analyze with computer support. We have
formalized them using the proof assistant Isabelle/HOL \citep{Isabelle}
together with the automated higher-order reasoners Leo-II \citep{C26},
Satallax \citep{brown2012satallax}, Metis 
\citep{Hurd03first-orderproof}, and Nitpick \citep{Nitpick}.   Our
formalizations\footnote{The   formalizations are available in the
subdirectories \url{Anderson}, \url{Hájek} and \url{Bjordal} at \url{gi
thub.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/
}.} employ the embedding of higher-order modal logic (HOML) in
classical higher-order logic (HOL) as introduced in previous work
\citep{C40,J30,J23}. We explored the effect of different domain
conditions on the provability of lemmas, theorems and even axioms.
This was motivated by a controversy between Hájek and Anderson
regarding the redundancy of some axioms in Anderson's emendation. In
\emph{constant domain semantics}, the individual domains are the same
in all possible worlds. In \emph{varying domain semantics}, the
domains may vary from world to world. This variation is technically
encoded with the help of an existence relation expressing which
individuals actually exist in each world. Quantifiers are then
uniformly formalized as \emph{actualistic quantifiers} (i.e. guarded by
the existence relation). Our main results are summarized here.

@@

For all emendations and variants discussed here, the axioms and definitions have been shown to be consistent and 
not to entail modal collapse.

@@

The redundancy of A4 and A5 is particularly controversial. Magari \cite{Magari1988} claimed that A4 and A5 are superfluous\footnote{
  An axiom $A$ is superfluous w.r.t. a set of axioms $S$ iff T3 is entailed by $S \setminus \{ A \}$.
}, arguing that T3 is true in all models of the other axioms and definitions by Gödel. \citet[p.~5-6]{Hajek_Magari_and_others_1996} investigated this further, refuted Magari's claim, but claimed that A4 and A5 are indeed superfluous under the assumption of an additional axiom (PEP, pointwise equality for positiveness), which Magari probably assumed silently. Moreover, Hájek \citet[p.~2]{Hajek_Magari_and_others_1996} cites his earlier work\footnote{Although \citep{Hajek_der_Mathematiker_2001} precedes \citep{Hajek_Magari_and_others_1996} in writing, it was published only 5 years later, in German.} \citep{Hajek_der_Mathematiker_2001}, where he claims (in Theorem 5.3) that for Anderson's emended theory \citep{anderson90:_some_emend_of_goedel_ontol_proof}, A4 and A5 are not only superfluous, but also redundant. \citet[footnote 1 in p.~1]{AndersonGettings}, in a footnote, rebutted Hájek's claim, arguing that the redundancy of A4 and A5 holds only under constant domain semantics, while Anderson's emended theory ought to be taken under Cocchiarella's semantics \citep{Cocchiarella} (a varying domain semantics). Our results show that Hájek was originally right, under both constant and varying domain semantics.

@@

Additionally, Anderson's 14th footnote \citep[footnote 14]{anderson90:_some_emend_of_goedel_ontol_proof} remarks that only the quantifiers in T3' and in A:D2 need to be interpreted as actualistic quantifiers, while others may be taken as possibilistic quantifiers. Our computer-assisted study of this mixed variant shows that A4 is still redundant, but A5 becomes independent (hence not redundant). Unfortunately, T3 then becomes unprovable.

The controversy over the superfluousness of A4 and A5 indicates a trend to reduce the ontological argument to its bare essentials. In this regard, 
already \citet[p.~7]{anderson90:_some_emend_of_goedel_ontol_proof} indicates that, by taking a notion of \emph{defective} as primitive and defining the notion of \emph{positive} upon it, axioms A:A1, A2 and A4 become derivable. These claims have been confirmed by the automated theorem provers. Within the same trend, the alternative proposed by \citet{bjordal99} (cf. Appendix \ref{apx:Bjordal}) achieves a high level of minimality. He takes the property of being God-like as a primitive and defines (B:D1) the positive properties as those properties necessarily possessed by every God-like being. He then briefly indicates (B:L1) that B:D1 is logically equivalent, under modal logic \SFour, to the conjunction of D, A2, A3 and A4. This has been confirmed in the computer-assisted formalization: A2 and A3 can be quickly automatically derived in logic \K. A4 can be proved in logic \KT (i.e. assuming reflexivity of the accessibility relation). Proving D1 is possible in logic \KFour. Conversely, the proof that B:D1 is entailed by D1, A2, A3 and A4 is possible already in logic \K. The provers also show that theorem T3 follows from B:D1, B:D2, B:D3, A:A1 and A5 already in logic \KB. Bjordal's last paragraph briefly mentions Hájek's ideas about the superfluousness of A5 and claims that it is possible, with certain additional presuppositions, to eliminate A5 from his theory as well. Without any additional presupposition, the automated reasoners show that A5 is actually not superfluous. All these results hold for both constant and varying domain semantics.

Using our approach, the formalization and (partly) automated
analysis of several variants of G\"odel's ontological argument has been surprisingly
straightforward. The provers not only confirmed many claimed results, but also exposed a few mistakes and novel insights in a long standing controversy. 
We believe the technology employed in this work is ready to be 
fruitfully adopted in larger scale by philosophers.

% ToDo: \\
% 3 - Fix Hajek's proofs in the appendices. (they are just copy-pasted from Anderson's so far)
% 4 - Check apostrophes in axiom names that depend on different definitions.
% 5 - Add subscripts to alternative definitions (e.g. G_A)
% 6 - Fix citet redundancy
% 7 - Add apendix for Anderson's simplification

@@

@@

% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
In Bjordal's emendation $G$ (God-like) is taken as primitive and $P$ (Positive) is defined (cf. definition D).
\begin{itemize}
\item[B:D1] A formulas $\phi$ is positive iff it is necessarily the case
  that anything which is God-like has the property $\phi$.
  $$\hol{P(\phi) \equiv \nec \allq x (G(x) \imp \phi(x))}$$ 
\item[B:L1] D is logically equivalent in S4 with the union of G\"odel's definition D1 and axioms A2, A3 and A4.
  $$\hol{D \biimp D1 \wedge A2 \wedge A3 \wedge A4}$$
  The proof splits into the two implication directions B:L1$^\rightarrow$ and B:L1$^\leftarrow$. B:L1$^\rightarrow$ can be further split into four single steps.
\item[B:D2] $\phi$ is a maximal composite of object $x$'s positive properties iff $x$ has $\phi$ and $\phi$ is positive and all positive properties $\psi$ which $x$ has are such that is necessarily the case that all objects which have $\phi$ also have $\psi$.
  $$\hol{MCP(\phi,x) \equiv (\phi(x) \wedge P(\phi)) \wedge \allq \psi ((\psi(x) \wedge P(\psi)) \imp \nec \allq y (\phi(y) \imp \psi(y)))}$$
\item[B:D3] $x$ has the $N$-property iff x is such that if $\phi$ is a maximal composite of $x$'s positive properties then it is necessary that some object $y$ has the property $\phi$.
  $$\hol{N(x) \equiv \allq \phi (MCP(\phi,x) \imp \nec \allq y \phi(y))}$$
\item[A:A1] If a property is positive, its negation is not positive:
  $$\hol{\allq \varphi [P(\varphi) \imp \neg P(\neg \varphi)]}$$ 
\item[A5'] The $N$-property is positive.
 $$\hol{P(N)}$$
\item[T3'] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Frode's Alternative \cite{ToDo:Frode}.\label{fig:frode}} 
% \end{figure}
